---
title: "Probability theory 2"
output: 
  learnr::tutorial:
    progressive: true   
    allow_skip: true
runtime: shiny_prerendered
author: Shyam Gopalakrishnan
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(viridis)
knitr::opts_chunk$set(echo = TRUE)
```

## Today's programme

Similar to yesterday's tutorial, we will use various R functions to play around with continuous random variables.
Specifically, we will cover the following topics

-   Uniform distribution
-   Exponential distribution
-   Normal distribution
-   Playing with central limit theorem

For all of these distributions, we will cover sample from them, compute expectations and variance.

## Uniform distribution

We will start today's exploration of the continuous distributions with the Uniform distribution.
Let us start by sampling a single number from a uniform distribution between 0 and 1.
We will use the `runif` function to do this.

```{r}
runif(n=1, min=0, max=1)
```

Try adjusting the min and max options to change the support of the distribution to -1 and 1.

```{r unif1, exercise = TRUE}

```

Try copying the original code multiple types, then run it.
Do you always get the same result?
Do you ever get the same result?

```{r unif2, exercise = TRUE}

```

The first input parameter (n) determines the number of samples we are asking for.
Try increasing it to 50 samples, and see if you get a repeat observation.
Would you expect a repeat observation?

```{r unif3, exercise = TRUE}
```

Let us try and compute the pdf of the Uniform distribution on (0, 1) at the value of 0.3.

```{r}
dunif(0.3, 0, 1)
```

What did you get for `f(x=0.3)` for the uniform distribution?
Remember that you can always look up help for R functions using the `?` operator, e.g. `?dunif`.

Try to change the value at which we are computing the pdf, across various choices of x between 0 and 1.
What do you get for pdf at these points?
What would you expect?
Try this for the distribution on (-1, 1)?
What would you expect happens to the value of f(x=0.3)?
Play around with the range of the distribution, and see what happens to the pdf: f(x)?

```{r unif4, exercise = TRUE}

```

Similar to the `runif` and `dunif` function, we have the `punif` and `qunif` functions that give you the CDF and quantile function respectively.
CDF we already know, but what is the quantile function?
Remember that CDF, F(x) = P(X \<= x), so `punif` gives the value F(x) for a given x.
The quantile function is the inverse of the CDF, so given a value b, it finds a value x, such that F(x) = b.
Let us use these to find the cdf at the point 0.79.
Also, find the value of x, such that the cdf is 0.35.

```{r}
punif(0.79)
qunif(0.35)
```

Try the same values, but with the range now being -1, 1.

```{r unif5, exercise = TRUE}

```

Let us now use random sampling to get an estimate of the mean of the random variable with Uniform distribution on (0,1).
For starters, let us use 100 samples.

```{r}
mean(runif(n=100))
```

Try it with 1000, 10000 and 100000 samples.
What happens to the mean estimate?
Try the same also for the variance.
How close do these get to the theoretical values for the uniform distribution?

```{r}

```

The change in the estimates of the mean and the variance should tell us the power of increasing the sample size.
For each sample size, do this 100 times, and plot the results using a box plot to see how the accuracy of the mean estimate changes with sample size.

## Exponential distribution

The second continuous distribution that we talked about was the exponential distribution.
Just like we did with the uniform distribution, we have the functions `rexp`, `pexp`, `qexp` and `dexp` for the exponential distribution that will allow us to sample, get the CDF, quantile function and pdf for this distribution.
Remember that the exponential distribution has only one parameter $\lambda$ that is the rate parameter.

As we did for the uniform distribution, let us start by sampling from the exponential distribution with rate 5.

```{r}
rexp(1, 5)
```

Now let us sample 100 samples from this distribution, and plot a histogram of the values.

```{r exp1, exercise = TRUE}

```

What does the histogram look like?
Repeat the same experiment with rate values of 1 and 10.
How do the histograms change?

```{r exp2, exercise = TRUE}

```

We can also use the `dexp` function to get the plot for the PDF of this distribution.

```{r}
x = seq(0, 5, by=0.01)
exp_pdf = dexp(x, rate=5)
plot(x, exp_pdf, type="l", col="red", xlab="X", ylab="Exponential pdf")
```

You can also try to do the same for rates of 1 and 10, and compare them with the histograms you get.

```{r exp25, exercise = TRUE}

```

We will now use the sampling function to compute various statistics for the exponential distribution, and see how well they match up with what we know about the distribution from theory.
First let us set the rate parameter $\lambda = 5$.
Using this and a sample size of 1000, compute the probability that the random variable with this exponential distribution is less than 0.5, and then 0.2.

```{r}
exp_sample = rexp(1000, 5)
mean(exp_sample < 0.5)
mean(exp_sample > 0.2)
```

Let us now use the `pexp` function to verify if our estimates are correct.
Use the help to figure out what the `pexp` function inputs.

```{r exp3, exercise = TRUE}

```

For the exponential distribution, we know that the mean is given by $\lambda^{-1}$.
So using the sampling method, for a rate of 5, and a sample size of 100, 1000, and 10000, compute the mean of the random variable with this distribution.
I have the code for 100 samples.
Add the code for 1000 and 10000 samples.
Do you see the same effect as you did for the uniform distribution?

```{r exp4, exercise = TRUE}
exp_sample = rexp(100, 5)
mean(exp_sample)
var(exp_sample)
```

## Normal distribution

The normal distribution (Gaussian) is the most important distribution in probability theory and statistics - we will discuss more the why in the next section.
As with the other two distributions that we have already looked at today, the functions `rnorm`, `dnorm`, `pnorm` and `qnorm` let you compute the various aspects of the normal distribution.

Remember that the normal distribution is parameterized by two parameters, the mean $\mu$ and the standard deviation, $\sigma$.
Let us start with the distribution with a mean of 0 ($\mu=0$) and a standard deviation of 1 ($\sigma=1$).
Let us start by simulating 1000\
samples from this distribution, and plotting a histogram with them - to recover the bell curve.
Let us then compare it with the expected pdf for this normal between -5 and 5.

```{r}
norm_samples = rnorm(1000, 0, 1)
x = seq(-5,5,by=0.01)
norm_pdf = exp(-x**2)/sqrt(2*pi)
hist(norm_samples, border=NA, col="gray50", probability=TRUE, breaks=50)
lines(x, norm_pdf, type="l", col="firebrick")
```

Let us use the same simulation to see what proportion of the samples lie within 1 standard deviation of the mean.
Remember that the mean is 0, and standard deviation is 1, so we are looking for proportion of samples that lie in the window (-1, 1).
What number is this?
What about within 2 standard deviations (sd) of 0.
Code for 1 sd is given, try 2 and 3 sds away from the mean.

```{r norm1, exercise = TRUE}
norm_samples = rnorm(1000)
prop_1sd = mean(norm_samples > -1 & norm_samples < 1)
```

We know $P(a < X < b) = F(b) -F (a)$, so let us compute these proportions again from the CDF themselves.
Again code for the 1st case is given.

```{r norm15, exercise = TRUE}
prop_1sd_theoretical = pnorm(1) - pnorm(-1)
```

How do these compare to the proportion you got by counting the samples in that range?

Let us now try to do the same with a different mean and standard deviation - say mean of 10 and standard deviation of 5.
Generate a 1000 samples, and figure out proportion of samples that within 1, 2 and 3 sd of the mean.
What are these proportions?
How did they differ from the standard normal case.

```{r norm2, exercise = TRUE}

```

Let us use the same samples to compute the sample mean and sample variance of underlying distribution.

```{r norm3, exercise = TRUE}

```

Now let us compute the correlation between two normal distributions, one of which has mean 0 and sd 1, whereas the other has mean 10 and sd 5.

```{r norm4}
norm_samp_0_1 = rnorm(1000, 0, 1)
norm_samp_10_5 = rnorm(1000, 10, 5)
cor(norm_samp_0_1, norm_samp_10_5)
```

Use the formula from yesterday (and remember that variance is sd\^2) to compute the correlation theoretically, what would it be?

Additional exercise: \## Central limit theorem
